Optimizing Abstract Abstract Machines \vskip -1.5cm

### 2012-11-14 paper/paper.tex
10 pages.

\title{Optimizing Abstract Abstract Machines}

@@

Conceptually, the AAM approach calls for computing an analysis as a graph
exploration: (1) start with an initial state, and (2) compute the transitive
closure of the transition relation starting from that state.

@@

We have two choices for how to implement lazy non-determinism.

@@

We will explore both options here.
%
We will first explain the natural abstraction of tuples, and then we
will explain a less precise allocation strategy from uniform $k$-CFA
that we use for large compound data literals that we modified from the
implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.

@@

\item {\bf graphs}: counts the number of directed graphs with a
  distinguished root and \(k\) vertices, each having out-degree at
  most 2. It is written in a continuation-passing style and makes
  extensive use of higher-order procedures---it creates closures
  almost as often as it performs non-tail procedure calls--- (used by
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
\end{enumerate}

@@

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright and
Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
successor, the Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.  Many of our
  benchmarks are drawn from these papers.}  Unfortunately, these
analyses compute an inherently different and uncomparable form of
analysis.  Consequently, we have omitted a complete comparison with
these implementations,.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design.

@@

Constraint-based program analyses
(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
typically compute sets of abstract values for each program point.
These values approximate the values arising at run-time for each
program point.  Value sets are computed as the least solution to a set
of constraints (either inclusion or equality constraints).  The
constraints must be designed and proved as a sound approximation of
the semantics.  Efficient implementations of these kinds of analyses
often take the form of worklist-based graph algorithms for constraint
solving, and are thus quite different from the interpreter
implementation.  The approach thus requires effort in constraint
system design and implementation, and the resulting system require
verification effort to prove the constraint system is sound and that
the implementation is correct.

@@

The AAM approach, and the state transition graphs it generates, encode
temporal properties not found in classical constraint-based analyses
for higher-order programs.
%
These analyses (ultimately) compute judgments on program terms and
contexts, e.g., at expression $e$, variable $x$ may have value $v$.
%
The judgments do not relate the order in which expressions and context
may be evaluated in a program, e.g., it has nothing
to say with regard to question like, ``Do we always evaluate $e_1$
before $e_2$?'' or ``Is it always the case that a file handle is
opened, read and then closed in that order?''
%
The state transition graphs can answer these kinds of queries, but
this does not come for free: respecting temporal order imposes an
order in which states and terms may be evaluated \emph{during} the
analysis.

@@

Abstract machines are not only a good model for rapid analysis
development, they also can be systematically developed into efficient
algorithms.

@@

\balance
\bibliographystyle{plain}
\bibliography{local,bibliography}

>>>>

\title{Optimizing Abstract Abstract Machines
\vskip -1.5cm}

@@

Conceptually, the AAM approach calls for computing an analysis as a graph
exploration: (1) start with an initial state, and (2) compute the transitive
closure of the transition relation from that state.

@@

\noindent
We have two choices for how to implement lazy non-determinism.

@@

We explore both options here.
%
We will first explain the natural abstraction of tuples, and then we
will explain a less precise allocation strategy from uniform $k$-CFA
that we use for large compound data literals that we modified from the
implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.

@@

\item {\bf graphs}: counts the number of directed graphs with a
  distinguished root and \(k\) vertices, each having out-degree at
  most 2. It is written in a continuation-passing style and makes
  extensive use of higher-order procedures---it creates closures
  almost as often as it performs non-tail procedure calls (used by
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
\end{enumerate}

@@

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright and
Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
successor, the Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.}  Unfortunately, these
analyses compute an inherently different and uncomparable form of
analysis.  Consequently, we have omitted a complete comparison with
these implementations,.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design.

@@

Constraint-based program analyses
(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
typically compute sets of abstract values for each program point.
These values approximate values arising at run-time for each
program point.  Value sets are computed as the least solution to a set
of (inclusion or equality) constraints.  The
constraints must be designed and proved as a sound approximation of
the semantics.  Efficient implementations of these kinds of analyses
often take the form of worklist-based graph algorithms for constraint
solving, and are thus quite different from the interpreter
implementation.  The approach thus requires effort in constraint
system design and implementation, and the resulting system require
verification effort to prove the constraint system is sound and that
the implementation is correct.

@@

The AAM approach, and the state transition graphs it generates, encodes
temporal properties not found in classical constraint-based analyses
for higher-order programs.
%
These analyses (ultimately) compute judgments on program terms and
contexts, e.g., at expression $e$, variable $x$ may have value $v$.
%
The judgments do not relate the order in which expressions and context
may be evaluated in a program, e.g., it has nothing
to say with regard to question like, ``Do we always evaluate $e_1$
before $e_2$?'' or ``Is it always the case that a file handle is
opened, read and then closed in that order?''
%
The state transition graphs can answer these kinds of queries, but
this does not come for free: respecting temporal order imposes an
order in which states and terms may be evaluated \emph{during} the
analysis.

@@

Abstract machines are not only a good model for rapid analysis
development, they can be systematically developed into efficient
algorithms.

@@

\balance
\bibliographystyle{plain}
\input{paper.bbl}
%\bibliography{local,bibliography}
